Copulas and their potential for ecology

### 2019-12-23 Paper.Rmd
suppmat notation table need to fix, created with NA as of now
minor corrections made while checking Shyas conversion to Word

\noindent \emph{The number of words in the abstract:} 253

@@

All branches of ecology 
study relationships among biological variables and relationships between 
environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description the relationship. 
For instance, datasets showing different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of association that can
have ecological significance, as discussed below, but that is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

We give a brief introduction to copulas, focusing, for simplicity, on bivariate 
copulas and on concepts needed
for the rest of the paper. We assume familiarity with the basic 
language of probability distributions and random variables.
See @nelsen2006_copula, @Genest2007, @joe2014_dependence, and
@MaiScherer2017 for general background on copulas, 
and see also @Anderson2018, part of 
which is a short introduction to 
copulas for ecologists. A bivariate *copula* can be defined as a bivariate cdf 
with both margins uniform on $(0,1)$ (@joe2014_dependence, p. 7). 
This will be the 
cdf of a bivariate random variable with uniform marginals, and 
the terminology *copula* is sometimes also applied to this random vector. 
A foundational theorem of
@sklar1959_theorem (see also @MaiScherer2017, p. 16) 
says that if $F$ is the cdf
of a random vector $(X,Y)$, with margins $F_X$ and $F_Y$, then there 
exists a copula $C$
such that for all $(x,y)$ in the Euclidean plane, $F(x,y)=C(F_X(x),F_Y(y))$.
The theorem also says $C$ is unique if $F_X$ and $F_Y$ are continuous.
Thus $C$ couples the bivariate cdf, $F$, with the 
cdfs of the marginals. 
Finally, Sklar's theorem says that if $D$ is any bivariate copula and 
$G_A$ and $G_B$ are univariate cdfs, then 
$D(G_A(x),G_B(y))$ is a valid cdf 
of some bivariate random variable.
The applications of this study fall within the case in which univariate marginal 
distributions have continuous, strictly 
monotonic cdfs, and this case is simpler. So we 
henceforth make such an assumption. Table SX provides a summary
of notation.
<!--***DAN: Shyamolina, once I email you an edited version of the
table you produced, please make a table in the sup mat Rmd and cause it to be
cited here in place of the SX above. Please place the table in such
a way that SX is replaced with S1,  since I believe this is now going
to be the first supplementary table we are citing.-->

@@

We fit 16 bivariate copula families, exhibiting a variety of 
lower- and upper-tail dependence characteristics, with bivariate datasets. The purpose
of using a large collection of families was to include a 
variety of alternative dependence structures to have a robust model
selection and multi-model inference procedure. For that purpose, it 
is not important for the reader to
understand the details of these copulas,
and, additionally, these copulas have been 
described in detail elsewhere [@joe1997multivariate; @Brechmann2013]. So 
we only briefly identify each family, say a few words about its tail 
dependence, and provide pictorial descriptions (Figs 
\ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8}). 
The pictorial descriptions can also be used to aid quick comparisons of copula
families when alternative families are being considered for future applications. 
We used several families that can exhibit positive lower-tail dependence 
(of strength depending on parameters) and 
zero upper-tail dependence: the Clayton, survival Gumbel, survival Joe
and survival BB6 families. We used several families that can exhibit positive
upper-tail dependence (strength depends on parameters) and zero lower-tail
dependence: the survival Clayton, Gumbel, Joe and BB6 families. We used
families that show zero upper- and lower-tail dependence: the normal
and Frank families. These families have pdfs symmetric about the line
$v=-u+1$. We also used several families that can show both upper- and
lower-tail dependence, in relative amounts depending on parameter
values: the BB1, survival BB1, BB7 and survival BB7 families. 
We also used the BB8 family, which shows zero lower-tail dependence,
and zero upper-tail dependence except for a boundary case for the parameters.
And we used the survival BB8 copula, which shows zero upper-tail dependence, 
and zero lower-tail dependence except for a boundary case.
"Survival" families are rotations of the copula with the similar name by
180 degrees. 
See @joe1997multivariate and @Brechmann2013 for details on all families. We used the implementations
of provided in the `VineCopula` package for the R programming
language. See Figs \ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8} 
for visual depictions of the pdfs of these copulas and how
pdfs and tail dependence are influenced by the parameters. See 
Table \ref{tab_copula_info} for a summary.

@@

row.names(tab_multivar_summary_ed)<-NULL
library(kableExtra)
library(dplyr)
library(tibble)
dt <- tibble(
            Description = c("1. Location pairs (excluding self comparisons)",
           "2. Number of non-independent pairs",
           "3. Number of pairs with non-normal copula as best fit",
           "4. Average of AIC differences (normal AIC minus min AIC) across location pairs",
           "5. Number of well-fitting location pairs",
           "6. 2.5$^{th}$ and 97.5$^{th}$ quantiles of model-avg. lower-tail dependence",
           "7. 2.5$^{th}$ and 97.5$^{th}$ quantiles of model-avg. upper-tail dependence",
           "8. Percent pairs showing stronger lower tail dependence",
           "9. Mean with 2.5$^{th}$ and 97.5$^{th}$ quantiles of cor$_{l}$ - cor${_u}$ across all location pair")
           )

@@

Our second mechanism is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
We consider a linear model, as well as 
two parameterizations of a nonlinear population model which includes density dependence
of population growth. For the linear model, let $E_i(t)$ 
again be an ecological variable, $i=1,2$.
We use AR(1) dynamics, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \varepsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\varepsilon_i(t)$ were 
standard-normal random variables that were independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is general. It could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.
The nonlinear population model was a stochastic, multi-habitat-patch version 
of the Ricker model, 
$P_i(t+1)=P_i(t)\exp \left[r\left(1-P_i(t)/K  \right)+\sigma\varepsilon_i(t)\right]$ for $i=1,2$, using $r=0.5$, 
$K=100$, $\varepsilon_i(t)$
as above, and $\sigma=0.1$ or 
$\sigma=1$ in different simulations. 
When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, where the model equation can be well approximated by a linear equation,
and the nonlinearities of the model therefore have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear because the stochastic 
component of the model causes populations to stray far from the carrying capacity. 
We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when strong, it brings the nonlinearities of the model into play.

@@

<!--referee-requested para comparing to regression with heteroskedastic error term-->
Non-copula methods may be useful under some circumstances for exploring
asymmetries of tail association, but copula approaches have theoretical
and practical advantages. It was suggested to us that asymmetries of
tail association can also be explored using a regression 
model with heteroskedastic error term, $y=mx+b+\varepsilon(x)$. Whereas this 
approach seems capable of yielding insight for some datasets, it implicitly assumes
a particular causal relationship between $x$ and $y$, an untenable assumption for 
many applications. Furthermore, no equivalent
of Sklar's theorem exists under this approach, so information on marginal 
distributions cannot be separated from association information, and the two types
of information will both influence regression results. Sklar's theorem
means that the copula associated with two variables contains all and only the information
about the association. Such statements of 
mathematical completeness will not be avilable for other methods.

>>>>

\noindent \emph{The number of words in the abstract:} 241

@@

All branches of ecology 
study relationships among biological variables and relationships between 
environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of the relationship. 
For instance, datasets showing different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of association that can
have ecological significance, as discussed below, but that is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

We give a brief introduction to copulas, focusing, for simplicity, on bivariate 
copulas and on concepts needed
for the rest of the paper. We assume familiarity with the basic 
language of probability distributions and random variables.
See @nelsen2006_copula, @Genest2007, @joe2014_dependence, and
@MaiScherer2017 for general background on copulas, 
and see also @Anderson2018, part of 
which is a short introduction to 
copulas for ecologists. A bivariate *copula* can be defined as a bivariate cdf 
with both margins uniform on $(0,1)$ (@joe2014_dependence, p. 7). 
This will be the 
cdf of a bivariate random variable with uniform marginals, and 
the terminology *copula* is sometimes also applied to this random vector. 
A foundational theorem of
@sklar1959_theorem (see also @MaiScherer2017, p. 16) 
says that if $F$ is the cdf
of a random vector $(X,Y)$, with margins $F_X$ and $F_Y$, then there 
exists a copula $C$
such that for all $(x,y)$ in the Euclidean plane, $F(x,y)=C(F_X(x),F_Y(y))$.
The theorem also says $C$ is unique if $F_X$ and $F_Y$ are continuous.
Thus $C$ couples the bivariate cdf, $F$, with the 
cdfs of the marginals. 
Finally, Sklar's theorem says that if $D$ is any bivariate copula and 
$G_A$ and $G_B$ are univariate cdfs, then 
$D(G_A(x),G_B(y))$ is a valid cdf 
of some bivariate random variable.
The applications of this study fall within the case in which univariate marginal 
distributions have continuous, strictly 
monotonic cdfs, and this case is simpler. So we 
henceforth make such an assumption. Table \ref{SM-tab_summary_notation} provides a summary
of notation.
<!--***DAN: Shyamolina, once I email you an edited version of the
table you produced, please make a table in the sup mat Rmd and cause it to be
cited here in place of the SX above. Please place the table in such
a way that SX is replaced with S1,  since I believe this is now going
to be the first supplementary table we are citing.-->

@@

We fit 16 bivariate copula families, exhibiting a variety of 
lower- and upper-tail dependence characteristics, with bivariate datasets. The purpose
of using a large collection of families was to include a 
variety of alternative dependence structures to have a robust model
selection and multi-model inference procedure. For that purpose, it 
is not important for the reader to
understand the details of these copulas,
and, additionally, these copulas have been 
described in detail elsewhere [@joe1997multivariate; @Brechmann2013]. So 
we only briefly identify each family, say a few words about its tail 
dependence, and provide pictorial descriptions (Figs 
\ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8}). 
The pictorial descriptions can also be used to aid quick comparisons of copula
families when alternative families are being considered for future applications. 
We used several families that can exhibit positive lower-tail dependence 
(of strength depending on parameters) and 
zero upper-tail dependence: the Clayton, survival Gumbel, survival Joe
and survival BB6 families. We used several families that can exhibit positive
upper-tail dependence (strength depends on parameters) and zero lower-tail
dependence: the survival Clayton, Gumbel, Joe and BB6 families. We used
families that show zero upper- and lower-tail dependence: the normal
and Frank families. These families have pdfs symmetric about the line
$v=-u+1$. We also used several families that can show both upper- and
lower-tail dependence, in relative amounts depending on parameter
values: the BB1, survival BB1, BB7 and survival BB7 families. 
We also used the BB8 family, which shows zero lower-tail dependence,
and zero upper-tail dependence except for a boundary case for the parameters.
And we used the survival BB8 copula, which shows zero upper-tail dependence, 
and zero lower-tail dependence except for a boundary case.
"Survival" families are rotations of the copula with the similar name by
180 degrees. 
See @joe1997multivariate and @Brechmann2013 for details on all families. We used the implementations
provided in the `VineCopula` package for the R programming
language. See Figs \ref{figped_N}-\ref{figped_BB1}, \ref{SM-figped_SC} and \ref{SM-figped_G}-\ref{SM-figped_SBB8} 
for visual depictions of the pdfs of these copulas and how
pdfs and tail dependence are influenced by the parameters. See 
Table \ref{tab_copula_info} for a summary.

@@

row.names(tab_multivar_summary_ed)<-NULL
library(kableExtra)
library(dplyr)
library(tibble)
dt <- tibble(
            Description = c("1. Location pairs (excluding self comparisons)",
           "2. Number of non-independent pairs",
           "3. Number of pairs with non-normal copula as best fit",
           "4. Average of AIC differences (normal AIC minus min AIC) across location pairs",
           "5. Number of well-fitting location pairs",
           "6. 2.5$^{th}$ and 97.5$^{th}$ quantiles of model-avg. lower-tail dependence",
           "7. 2.5$^{th}$ and 97.5$^{th}$ quantiles of model-avg. upper-tail dependence",
           "8. Percent pairs showing stronger lower tail dependence",
           "9. Mean with 2.5$^{th}$ and 97.5$^{th}$ quantiles of cor$_{l}$ - cor${_u}$ across all location pairs")
           )

@@

Our second mechanism is an extension of the well-known Moran
effect, and was summarized conceptually in the Introduction. 
We consider a linear model, as well as 
two parameterizations of a nonlinear population model which includes density dependence
of population growth. For the linear model, let $E_i(t)$ 
again be an ecological variable, $i=1,2$.
We use AR(1) dynamics, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \varepsilon_i(t)$, with $\beta=0.5$. 
The environmental noises $\varepsilon_i(t)$ were 
standard-normal random variables that were independent for distinct times, $t$, but 
exhibited different kinds of 
dependence across locations in different simulations (see below).
The variable $E$ is general. It could represent deviations
of a population density from a carrying capacity, deviations of
total plant community biomass from an average value,
flux of a biogeochemical variable such as methane, or other quantities.
The nonlinear population model was a stochastic, multi-habitat-patch version 
of the Ricker model, 
$P_i(t+1)=P_i(t)\exp \left[r\left(1-P_i(t)/K  \right)+\sigma\varepsilon_i(t)\right]$ for $i=1,2$, using $r=0.5$, 
$K=100$, $\varepsilon_i(t)$
as above, and $\sigma=0.1$ or 
$\sigma=1$ in different simulations. 
When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, where the model equation can be well approximated by a linear equation,
and the nonlinearities of the model therefore have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear because the stochastic 
component of the model causes populations to stray far from the carrying capacity. 
We refer to these as
the weak-noise and strong-noise cases, though the importance of the noise here
is that, when strong, it brings the nonlinearities of the model into play.

@@

<!--referee-requested para comparing to regression with heteroskedastic error term-->
Non-copula methods may be useful under some circumstances for exploring
asymmetries of tail association, but copula approaches have theoretical
and practical advantages. It was suggested to us that asymmetries of
tail association can also be explored using a regression 
model with heteroskedastic error term, $y=mx+b+\varepsilon(x)$. Whereas this 
approach seems capable of yielding insight for some datasets, it implicitly assumes
a particular causal relationship between $x$ and $y$, an untenable assumption for 
many applications. Furthermore, no equivalent
of Sklar's theorem exists under this approach, so information on marginal 
distributions cannot be separated from association information, and the two types
of information will both influence regression results. Sklar's theorem
means that the copula associated with two variables contains all and only the information
about the association. Such statements of 
mathematical completeness will not be available for other methods.
